\blue{...Numerical investigation of viscous flow instabilities through heterogeneous porous media using adaptive and unstructured meshes under different mobility scenarios...}

### 2017-07-05 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
updates
do not rush kostas

\medskip
Fluid flow dynamics through heterogeneous porous media are sensibly more complex than in homogeneous media and strongly depend on problem properties. Spatial variation in geological formations occurs in all length-scales, where heterogeneity characteristics in small length-scales (\ie pore) are statistically embedded into permeability (absolute and relative) and porosity parameters. Such multi-scale heterogeneity induces preferential flow pathways and plays a significant role in the onset of fluid instabilities as it triggers fingers formation and their accelerated growth \citep[see][]{ewing_1989,tchelepi_1994}.

In order to qualitatively investigate the impact of heterogeneity (here represented by changing in the permeability field), numerical simulations were conducted using a prescribed permeability distribution and viscosity ratios of 10 and 150. Absolute permeability ranging from 1.0$\times$10$^{-12}$ to 5.0$\times$10$^{-10}$ cm$^{2}$ (\ie 0.1 $\le$ {\bf K} $\le$ 50 milidarcy) was used in the simulations, which were performed with the same geometry,  mesh resolution, boundary and initial conditions as in previous homogeneous cases.

Figures~\ref{fig:HeleShawHeter_VR10} and ~\ref{fig:HeleShawHeter_VR150} show formation of multi-scale elongations with continuous growth and coalescence of dendritic finger branching in simulations performed with viscosity ratio of 10 and 150, respectively. In both simulations, maximum fingers width of approximately 0.44-0.56 ({\it VR}=10) and 0.24-0.44 cm were found, whereas estimated values (based on Eqn.~\ref{eqn:guaneqn}) were 0.25-4.50 ({\it VR}=10) and 0.88-3.00 cm (Fig.~\ref{fig:homoheleshaw_VN10_VN150} c and d, Table~\ref{table:Heleshaw}). Such calculated ranges \citep[\ie theoretical values based on analytic solutions due to][]{mclean_1981} clearly overestimate fingers dimensions as the mathematical formulation (and therefore Eqn.~\ref{eqn:guaneqn}) assumes incompressible flows in homogeneous porous medium, \ie it does not take into account any spatial variability of heterogeneity.

>>>>

\medskip
Fluid flow dynamics through heterogeneous porous media are sensibly more complex than in homogeneous media and strongly depend on problem properties. Spatial variation in geological formations occurs in all length-scales, where characteristics of heterogeneities in small (pore) length-scales are statistically embedded into permeability (absolute and relative) and porosity parameters. Such multi-scale heterogeneity induces preferential flow pathways and plays a significant role in the onset of fluid instabilities as it triggers fingers formation and their accelerated growth \citep[see][]{ewing_1989,tchelepi_1994}.

In order to qualitatively investigate the impact of heterogeneity (here represented by changing in the permeability field), numerical simulations were conducted using a prescribed permeability distribution and viscosity ratios of 10 and 150. Absolute permeability ranging from 1.0$\times$10$^{-12}$ to 5.0$\times$10$^{-10}$ cm$^{2}$ (\ie 0.1 $\le$ {\bf K} $\le$ 50 milidarcy) was used in the simulations, which were performed with the same mesh resolution, boundary and initial conditions as in previous homogeneous cases. Figures~\ref{fig:HeleShawHeter_VR10} and ~\ref{fig:HeleShawHeter_VR150} show formation of multi-scale elongations with continuous growth and coalescence of dendritic finger branching in simulations performed with viscosity ratio of 10 and 150, respectively. In both simulations, maximum fingers width of approximately 0.44-0.56 ({\it VR}=10) and 0.24-0.44 cm were found, whereas estimated values (based on Eqn.~\ref{eqn:guaneqn}) were 0.25-4.50 ({\it VR}=10) and 0.88-3.00 cm (Fig.~\ref{fig:homoheleshaw_VN10_VN150} c and d, Table~\ref{table:Heleshaw}). Such calculated \citep[\ie theoretical based on simplified mathematical formulation due to][]{mclean_1981} ranges clearly overestimate fingers dimensions as the mathematical formulation (and therefore Eqn.~\ref{eqn:guaneqn}) does not take into account spatial variability of heterogeneity.

### 2017-07-05 Papers/ViscousInstabilities/article_figure1.tex
do not rush kostas

%%%
%%% FIGURE XXXXXX
%%%
\begin{landscape}
  \begin{figure}[ht]
  \vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=.65\textwidth, height=.46\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D200Hb} 
            \includegraphics[width=.45\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D500Hb}
            \includegraphics[width=.45\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D800Hb} }
      \hbox{\hspace{2.0cm} (a) t=0.10s \hspace{4.cm} (b) t=0.24ss \hspace{4.0cm} (c) t=0.37s}
      \vspace{0.5cm}
      \hbox{\hspace{.5cm} \includegraphics[width=.45\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D1500Hb}
            \includegraphics[width=.45\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D5000Hb}
            \includegraphics[width=.45\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D8000Hb} }
      \hbox{\hspace{2.5cm} (d) t=0.59s \hspace{4.cm} (e) t= 1.61s\hspace{3.0cm} (f) t=2.50s }}
\caption{Simulated flow in a modified Hele-Shaw cell with {\it VR}=150: snapshots of saturation profile during 2.50 seconds of simulation. Permeability distribution used in this simulation was the same as shown in Fig.~\ref{fig:HeleShawHeter_VR10}a. The domain contains 26313 \PN[1]{2} element-pairs.}
\label{fig:HeleShawHeter_VR150}
\end{figure}
\end{landscape}
\clearpage

%%%%
%%%%  FIGURE
%%%%
\begin{figure}[ht] 
\vbox{
\hbox{\includegraphics[width=.5\textwidth]{./Pics1/Saffman_homogeneous_VR10/ST_Homog_VR10_D2201_W2b}
       \includegraphics[width=.5\textwidth]{./Pics1/Saffman_homogeneous_VR150/ST_Homog_VR150_D5003_W2b}}
\hbox{\hspace{0.5cm} (a) {\it VR}=10 (homogeneous) \hspace{2.cm} (b) {\it VR}=150 (homogeneous) }
\vspace{0.5cm}
\hbox{\includegraphics[width=.5\textwidth]{./Pics1/Saffman_heterogeneous_VR10/ST_Heterog_VR10_D11000_W2b}
       \includegraphics[width=.5\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D6500_W2b}}
\hbox{\hspace{0.5cm} (c) {\it VR}=10 (heterogeneous) \hspace{2.cm} (d) {\it VR}=150 (heterogeneous) }
}
\caption{Isosurfaces of simulated flows in Hele-Shaw cells with viscosity ratios of 10 (a and c) and 150 (b and d). Top and bottom rows describe simulations performed with constant $\left(\text{\ie homogeneous with {\bf K}= 10}^{-10}\text{cm}^{2}\right)$ and randomly distributed (\ie heterogeneous, Fig.~\ref{fig:HeleShawHeter_VR10}a) permeabilities. Width of largest fingers for homogeneous cases are approximately 0.70 and 0.90cm ({\it VR}=10 and {\it VR}=150, respectively), whereas for heterogeneous cases are 0.56 and 0.40cm. Results for homogeneous cases are in good agreement with values obtained from \citet{guan_2003}'s analytic solution.}
\label{fig:homoheleshaw_VN10_VN150}
\end{figure}

>>>>

%%%
%%% FIGURE XXXXXX
%%%
\begin{landscape}
  \begin{figure}[ht]
  \vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=.65\textwidth, height=.46\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D200Hb.pdf} 
            \includegraphics[width=.45\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D500Hb.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D800Hb.pdf} }
      \hbox{\hspace{2.0cm} (a) t=0.10s \hspace{4.cm} (b) t=0.24ss \hspace{4.0cm} (c) t=0.37s}
      \vspace{0.5cm}
      \hbox{\hspace{.5cm} \includegraphics[width=.45\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D1500Hb.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D5000Hb.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D8000Hb.pdf} }
      \hbox{\hspace{2.5cm} (d) t=0.59s \hspace{4.cm} (e) t= 1.61s\hspace{3.0cm} (f) t=2.50s }}
\caption{Simulated flow in a modified Hele-Shaw cell with {\it VR}=150: snapshots of saturation profile during 2.50 seconds of simulation. Permeability distribution used in this simulation was the same as shown in Fig.~\ref{fig:HeleShawHeter_VR10}a. The domain contains 26313 \PN[1]{2} element-pairs.}
\label{fig:HeleShawHeter_VR150}
\end{figure}
\end{landscape}
\clearpage

%%%%
%%%%  FIGURE
%%%%
\begin{figure}[ht]
\vbox{
\hbox{\includegraphics[width=.5\textwidth]{./Pics1/Saffman_homogeneous_VR10/ST_Homog_VR10_D2201_W2b.pdf}
       \includegraphics[width=.5\textwidth]{./Pics1/Saffman_homogeneous_VR150/ST_Homog_VR150_D5003_W2b.pdf}}
\hbox{\hspace{0.25cm} (a) {\it VR}=10 (homogeneous) \hspace{1.5cm} (b) {\it VR}=150 (homogeneous) }
\vspace{0.5cm}
\hbox{\includegraphics[width=.5\textwidth]{./Pics1/Saffman_heterogeneous_VR10/ST_Heterog_VR10_D11000_W2b.pdf}
       \includegraphics[width=.5\textwidth]{./Pics1/Saffman_heterogeneous_VR150/ST_Heterog_VR150_D6500_W2b.pdf}}
\hbox{\hspace{0.25cm} (c) {\it VR}=10 (heterogeneous) \hspace{1.5cm} (d) {\it VR}=150 (heterogeneous) }
}
\caption{Isosurfaces of simulated flows in Hele-Shaw cells with viscosity ratios of 10 (a and c) and 150 (b and d). Top and bottom rows describe simulations performed with constant \textit{\ie homogeneous with \textbf{K}}= $10^{-10}$ \textit{${cm}^{2}$} and randomly distributed (\ie heterogeneous, Fig.~\ref{fig:HeleShawHeter_VR10}a) permeabilities. Width of largest fingers for homogeneous cases are approximately $0.70$ and $0.90$cm ({\textit VR}=$10$ and {\textit VR}=$150$, respectively), whereas for heterogeneous cases are $0.56$ and $0.40$cm. Results for homogeneous cases are in good agreement with values obtained from \citet{guan_2003}'s analytic solution.}
\label{fig:homoheleshaw_VN10_VN150}
\end{figure}
